The imprint of global magma oceans on exoplanet demographics

### 2023-04-09 src/tex/ms.tex
6.6.2.

We currently see four potential lines of discriminating atmospheric signatures associated with super-runaway climates in the exoplanet population. (i) Detecting the atmospheric windows of water vapour in the near- to mid-infrared. In a super-runaway state the atmosphere is dominated by the opacity of water vapour, which has two prominent spectral features: one at 3.5--4.5 $\mu$m, and one between 8--20 $\mu$m \citep[e.g.][]{2021ApJ...919..130B}. This would require an instrument with capabilities in this wavelenght regime, for instance ELT METIS, JWST MIRI, or future missions such as LIFE \citep{2019A&A...621A.125B,2022A&A...664A..21Q}. Potentially the two-band filter capabilities of PLATO may offer insight into particularly pronounced spectral features in this super-runaway regime \citep{2020SSRv..216...98G}. (ii) Post-runaway planets may be detectable by O$_{3}$ absoprtion due to build-up of abiotic oxygen, left-over from photochemical dissociation and hydrogen loss \citep{2014ApJ...785L..20W,2015AsBio..15..119L}. (iii) Water loss in runaway greenhouse planets episodes would change the atmosphere's D/H isotopic ratio, akin to the enhancement in Venus' present-day atmosphere \citep{2019JGRE..124.2015K,2021JGRE..12606643K}. This may be detectable in high-resolution observations focusing on isotope-sensitive transitions \citep{2019AJ....158...26L,2019A&A...622A.139M}. (iv) Disequilibrium chemistry in tidally-locked super-runaway planets. It has been suggested that the atmospheric depth and the presence or absence of oceans on sub-Neptunes could be probed via the abundance of species that are photochemically destroyed in the upper atmosphere, and replenished from either thermochemical layers \citep{2021ApJ...914...38Y,2021ApJ...922L..27T} or at an ocean-atmosphere interface \citep{2019ApJ...887..231L,2021ApJ...921L...8H}. Atmospheric nitrogen and carbon compounds show strong sensitivity in their partitioning coefficients into magma \citep{2022E&PSL.59817847G,2022PSJ.....3...93B}. Therefore, it may be possible to discern the presence of an underlying magma ocean if the presence of an atmosphere on a rocky exoplanet can be confirmed. This has been suggested to be done via (v) eclipse photometry \citep{2019ApJ...886..141M,2019ApJ...886..140K} through the presence of a high albedo, which is expected to differ from the crystallized rock of a solidifed magma ocean \citep{2020ApJ...898..160E,Fortin2022}.

@@

\subsubsection{Baseline mass-radius relationship}
Our baseline mass-radius relationship assuming pure $\mathrm{MgSiO_3}$ interiors~\citep{Zeng2016} might not be representative of the rocky planet population.
However, while interior composition may change the magnitude of radius inflation and could introduce an offset to its demographic imprint, we do not expect a change of its structure.
The magnitude of the radius inflation effect is in fact expected to be larger for an Earth-like interior composition with an iron core-silicate mantle structure.\todo[inline]{TIM: do you have a reference for this statement?}
We thus consider our mass-radius relationship a conservative case and do not expect a negative impact on the detectability of the runaway greenhouse transition due to deviations from it.
Nevertheless, we performed a sanity check to assess the influence varying our baseline model on our results (see Appendix~\ref{app:MR_relation}), and found general agreement between different interior composition models.

The models in this study do not into account a predicted additional radius increase on the percent level due to a molten interior~\citep{Bower2019}, and we neglect any potential effects stemming from a deviating gas exchange between atmosphere and interior in runaway greenhouse planets~\citep{Ikoma2018}.
Future theoretical work that self-consistently models interior-atmosphere interactions is needed to take these effects into account~\citep{Turbet2019}.

>>>>

We currently see four potential lines of discriminating atmospheric signatures associated with super-runaway climates in the exoplanet population. (i) Detecting the atmospheric windows of water vapour in the near- to mid-infrared. In a super-runaway state the atmosphere is dominated by the opacity of water vapour, which has two prominent spectral features: one at 3.5--4.5 $\mu$m, and one between 8--20 $\mu$m \citep[e.g.][]{2021ApJ...919..130B}. This would require an instrument with capabilities in this wavelenght regime, for instance ELT METIS, JWST MIRI, or future missions such as LIFE \citep{2019A&A...621A.125B,2022A&A...664A..21Q,2022A&A...664A..22D}. Potentially the two-band filter capabilities of PLATO may offer insight into particularly pronounced spectral features in this super-runaway regime \citep{2020SSRv..216...98G}. (ii) Post-runaway planets may be detectable by O$_{3}$ absoprtion due to build-up of abiotic oxygen, left-over from photochemical dissociation and hydrogen loss \citep{2014ApJ...785L..20W,2015AsBio..15..119L}. (iii) Water loss in runaway greenhouse planets episodes would change the atmosphere's D/H isotopic ratio, akin to the enhancement in Venus' present-day atmosphere \citep{2019JGRE..124.2015K,2021JGRE..12606643K}. This may be detectable in high-resolution observations focusing on isotope-sensitive transitions \citep{2019AJ....158...26L,2019A&A...622A.139M}. (iv) Disequilibrium chemistry in tidally-locked super-runaway planets. It has been suggested that the atmospheric depth and the presence or absence of oceans on sub-Neptunes could be probed via the abundance of species that are photochemically destroyed in the upper atmosphere, and replenished from either thermochemical layers \citep{2021ApJ...914...38Y,2021ApJ...922L..27T} or at an ocean-atmosphere interface \citep{2019ApJ...887..231L,2021ApJ...921L...8H}. Atmospheric nitrogen and carbon compounds show strong sensitivity in their partitioning coefficients into magma \citep{2022E&PSL.59817847G,2022PSJ.....3...93B}. Therefore, it may be possible to discern the presence of an underlying magma ocean if the presence of an atmosphere on a rocky exoplanet can be confirmed. This has been suggested to be done via (v) eclipse photometry \citep{2019ApJ...886..141M,2019ApJ...886..140K} through the presence of a high albedo, which is expected to differ from the crystallized rock of a solidifed magma ocean \citep{2020ApJ...898..160E,Fortin2022}.

@@

\subsubsection{Baseline mass-radius relationship}
Our baseline mass-radius relationship assuming pure $\mathrm{MgSiO_3}$ interiors~\citep{Zeng2016} might not be representative of the rocky planet population.
However, while interior composition may change the magnitude of radius inflation and could introduce an offset to its demographic imprint, we do not expect a change of its structure.
The magnitude of the radius inflation effect is in fact expected to be larger for an Earth-like interior composition with an iron core-silicate mantle structure \citep[][, add Noack+ paper]{2016ApJ...819..127Z,2021JGRE..12606724B}.
We thus consider our mass-radius relationship a conservative case and do not expect a negative impact on the detectability of the runaway greenhouse transition due to deviations from it.
Nevertheless, we performed a sanity check to assess the influence varying our baseline model on our results (see Appendix~\ref{app:MR_relation}), and found general agreement between different interior composition models.

The models in this study do not into account a predicted additional radius increase on the percent level due to a molten interior~\citep{Bower2019}, and we neglect any potential effects stemming from a deviating gas exchange between atmosphere and interior in super-runaway greenhouse planets due to different redox conditions ~\citep{Ikoma2018,2021JGRE..12606711L,2022PSJ.....3...93B,2021SSRv..217...22G}.
Future theoretical work that self-consistently models interior-atmosphere interactions is needed to take these effects into account~\citep{Lichtenberg2022,2022ARA&A..60..159W}.
